Relationships in the extremes and their influence on competition and coexistence

### 2022-02-16 paper/Paper.Rnw
some edits to discussion

\underline{summary - include 1-2 sentences about misnomer}
We produced and applied a mathematical extension to MCT that, for the first time, 
considers ATAs for the coexistence of species and maintenance of biodiversity. 
Implementing new methods of MCT \citep{Ellner_2019, Ellner_2016}, we provided a further tertiary 
decomposition to MCT and decompose the storage effect into two new terms, 
$\Delta_i^{[EC]}$, the contribution of ATAs, and $\Delta_i^{[E \nat C]}$, the contribution of correlation \emph{per se}. 
The results of our application to a simple lottery model and 
to a competitive diatom model show that ATA contributions to storage effects
are non-negligible and meaningfully contribute to the coexistence of species and thus the maintenance of biodiversity. 
Our results reinforce that common methods of correlation fail to reflect and include 
important information on the structure of the relationship between variables and amends
coexistence to the list of ecological theory and phenomena that ATAs have been found to 
be important for \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Futhermore, though storage effects are typically described as the contribution
of covariance between drivers to coexistence, commonly referred to as \emph{EC covariance},
\citep{Chesson_2012, Ellner_2016, Ellner_2019}, we suggest that storage effects, instead, 
be described as the contribution of the \emph{relationship} between the drivers.
Measures of covariance do not fully capture all aspects of associativity throughout the bivariate 
distribution of the variables; thus, it may be misleading to describe the contribution of the whole 
EC (or other drivers') relationship as the contribution of EC (or other drivers') covariance.
Accordingly, we have separated the mechanisms of storage effect 
into contributions of ATAs and contributions of correlation \emph{per se}.

The example case exapanding upon coexistence decompositions of the \cite{Descamps_2005} chemostat experiment support our earlier findings that ATA contributions are not only non-zero, but are non-neglible and contribute meaningfully to coexistence.

@@

\underline{ATA specialization}
Not only is the contribution of ATA non-neglible but in some cases the presence of absence of ATA 
can mediate the result of competition between species, i.e., coexistence or extirpation. 
For some parameters, the contribution of ATAs was the driving mechanism of 
coexistence – the IGR would be negative without it – deeming the scenario a case of 
ATA rescue. For other parameter values ATA exclusion occurs where ATAs are the 
driving mechanism of the failure of coexistence and the lack of ATAs would result in 
a positive IGR. Parameter
subspaces of ATA exclusion and ATA rescue suggest a new kind of specialization among species or \emph{ATA specialists} - species which only persist in habitat patches where ATAs contribute to its coexistence. 
%I think ata specialists are a thing - if their competitor is ubiquitous, they can only live in places on the landscape 
%where there are atas. Could have a discussion para on it, if it does not make the results. Or both. BTW, ATA exclusion has a 
%similar thing going on - a species excluded by ATAs (and competition) can only live on the portions of the landscape where
%there are no ATAs. Could also describe the temporal way of it - the ATA specialist can only survive in a era/epoch for which there
%are ATAs in its local, otherwise it'll have to go dormant or move elsewhere.

@@

Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
Distinct increases in the magnitude of ATA contribution with increasing
environmental fluctuations, as shown in our results and
figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions evidently
become more variable.
As described in section \ref{section:introduction}, changes in the ATAs between environmental
variables may occur as the climate becomes more variable and extreme events become more 
expansive and extensive, causing relationships between separated events to become tighter at one end or another and thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible \emph{relationships} in the extremes; however, 
simpler changes in extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns of environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanims in other ways.
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

@@

\underline{E and vital rates}
%Discussion content: We should (in future) study tail associations between environmental variables, and how these might be changing. However, more than that, the variables E in the lottery model are not actual environmental variables but rather are the effects of environmental variables on the species of the model. Probably real environmental variables are actually related to vital rates in a non-linear and probably often threshold-like way. Threshold relationships can produce tail association, as we showed elsewhere. Changes in the means of true environmental variables may change where those distributions of environmental variables are compared to thresholds, which may change the nature of tail associations between the effects of those environmental variables.

\underline{the new effects are not "weak noise" effects}
if sigma really small, ATA would not matter?
%go back to emails
%Maybe have a paragraph about how Chesson's weak noise and other assumptions make it impossible to detect this new effect
%via that framework. It's only the less-assumption-reliant simulation approach of Ellner et al that makes it possible to 
%study this new term in the decomp. Need to study a bit to substantiate this properly, or get Steve Ellner help.

From proposal:

Classical analytic derivations of MCT using Taylor expansion approximations completely miss out
on any additions to coexistence beyond simple correlations. 
In fact, storage effects in analytic MCT is computationally equal to $\Delta^{[E \nat C]}$, 
the contributions of correlation per se. In newer, simulation based MCT, storage effect is 
the sum of the two new terms, $\Delta^{(EC)}=\Delta^{[E \nat C]}+\Delta^{[EC]}$, obscuring the effects of ATAs.

>>>>

\underline{summary - include 1-2 sentences about misnomer}
We produced and applied a mathematical extension to MCT that, for the first time, 
considers ATAs for the coexistence of species and maintenance of biodiversity. 
Implementing new methods of MCT \citep{Ellner_2019, Ellner_2016}, we provided a further tertiary 
decomposition to MCT and decompose the storage effect into two new terms, 
$\Delta_i^{[EC]}$, the contribution of ATAs, and $\Delta_i^{[E \nat C]}$, the contribution of correlation \emph{per se}. 
The results of our application to a simple lottery model and 
to a competitive diatom model show that ATA contributions to storage effects and to coexistence
are non-negligible and meaningfully contribute to the coexistence of species and thus the maintenance of biodiversity. 
Our results reinforce that common methods of correlation fail to reflect and include 
important information on the structure of the relationship between variables and amends
coexistence to the list of ecological theory and phenomena that ATAs have been found to 
be important for \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of covariance between drivers to coexistence, classically \emph{EC covariance},
\citep{Chesson_2012, Ellner_2016, Ellner_2019}, we suggest that storage effects, instead, 
be described as the contribution of the \emph{relationship} between the drivers.
Measures of covariance do not fully capture all aspects of associativity throughout the bivariate 
distribution of the variables; thus, it may be misleading to describe the contribution of the whole 
EC (or other drivers') relationship as the contribution of EC (or other drivers') covariance.
Accordingly, we have separated the mechanisms of storage effect 
into contributions of ATAs and contributions of correlation \emph{per se}.

%The example case exapanding upon coexistence decompositions of the \cite{Descamps_2005} chemostat experiment support our earlier findings that ATA contributions are not only non-zero, but are non-neglible and contribute meaningfully to coexistence.

@@

\underline{ATA specialization}
Not only is the contribution of ATA non-neglible but in some cases the presence or absence of ATAs 
can mediate the result of competition between species, i.e., coexistence or extirpation. 
For some parameters, the contribution of ATAs was the driving mechanism of 
coexistence – the IGR would be negative without it – deeming the scenario a case of 
ATA rescue. For other parameter values, ATA exclusion occurs where ATAs are the 
driving mechanism of the failure to coexist (exclusion) and the lack of ATAs would result in 
a positive IGR. Parameter
subspaces of ATA exclusion and ATA rescue suggest a new kind of specialization among species or \emph{ATA specialists} - species which only persist in habitat patches where ATAs contribute to its coexistence. 
%I think ata specialists are a thing - if their competitor is ubiquitous, they can only live in places on the landscape 
%where there are atas. Could have a discussion para on it, if it does not make the results. Or both. BTW, ATA exclusion has a 
%similar thing going on - a species excluded by ATAs (and competition) can only live on the portions of the landscape where
%there are no ATAs. Could also describe the temporal way of it - the ATA specialist can only survive in a era/epoch for which there
%are ATAs in its local, otherwise it'll have to go dormant or move elsewhere.

@@

Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
The coupling of the magnitude of ATA contribution and environmental variability, 
as shown in our results and figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), 
suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions evidently
become more variable.
As described in section \ref{section:introduction}, changes in the ATAs between environmental
variables may occur as the climate becomes more variable and extreme events become more 
expansive and extensive, causing relationships between separated events to become tighter at one end or another and thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible \emph{relationships} in the extremes; however, 
simpler changes in extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns between environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanims in other ways.
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

@@

\underline{E and vital rates}
We note that \emph{E}, in our investigation and in other investigations within the MCT 
framework, is not directly a measure of the environment, but is a measure of response to the 
environment (vital rates). Environment and environmental response may be related linearly
(like lottery model?); though probably real environmental variables are 
actually related to vital rates in a non-linear and often threshold-like way. 
Threshold relationships can produce tail association, as we showed elsewhere. 
Changes in the means of true environmental variables may change 
where those distributions of environmental variables are compared to thresholds, 
which may change the nature of tail associations between the effects of 
those environmental variables.
%Discussion content: We should (in future) study tail associations between environmental variables, and how these might be changing. However, more than that, the variables E in the lottery model are not actual environmental variables but rather are the effects of environmental variables on the species of the model.

\underline{the new effects are not "weak noise" effects}
if sigma really small, ATA would not matter?
%go back to emails
%Maybe have a paragraph about how Chesson's weak noise and other assumptions make it impossible to detect this new effect
%via that framework. It's only the less-assumption-reliant simulation approach of Ellner et al that makes it possible to 
%study this new term in the decomp. Need to study a bit to substantiate this properly, or get Steve Ellner help.
Classical analytic derivations of MCT using Taylor expansion approximations completely miss out
on any additions to coexistence beyond simple correlations. 
In fact, storage effects in analytic MCT is computationally equal to $\Delta^{[E \nat C]}$, 
the contributions of correlation \emph{per se}. 
In newer, simulation based MCT, storage effect is 
the sum of the two new terms, $\Delta^{(EC)}=\Delta^{[E \nat C]}+\Delta^{[EC]}$, 
obscuring the effects of ATAs.
